### 2015-05-13 whoop.tex
update to change sections to S symbol

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In Section~\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In Section~\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.

Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in Section~\ref{corral}.

@@

\medskip\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in Section~\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

@@

To achieve this, we first \emph{instrument} a thread template for lockset analysis (see Section~\ref{bg:lockset}).  Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.

@@

We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}. In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see Section~\ref{evaluation}) are indeed terminating.

@@

The simple concurrent programming model discussed in Section~\ref{sec:symbolicpairwise} is deliberately idealistic to describe our symbolic verification technique. In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.

@@

Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and Section~\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and Section~\ref{corral}).

@@

To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}. This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap. A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.\ pointers that cannot alias). We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in Section~\ref{corral}.

@@

\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see Section~\ref{sec:symbolicpairwise}). Such sets can be typically modeled in Boogie using maps. However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness. To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, which works as follows:

@@

Statically analyzing system-level software, such as drivers, requires to ``close'' the environment by abstracting away the low level implementation details. In this work, we developed a simple environmental model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in decreased precision and, thus, false positives. However, because we currently only focus on finding data races, we can get away with over-approximating a lot of the underlying Linux kernel functionality, without losing too much precision. Enriching our model with domain-specific knowledge to make it more precise is an ongoing manual effort, but requires Linux expertise. We argue that further increasing the precision is orthogonal to the contributions of this paper. Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{corral} and~\ref{evaluation}.

>>>>

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.

Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

\medskip\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

@@

To achieve this, we first \emph{instrument} a thread template for lockset analysis (see \S\ref{bg:lockset}).  Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.

@@

We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}. In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are indeed terminating.

@@

The simple concurrent programming model discussed in \S\ref{sec:symbolicpairwise} is deliberately idealistic to describe our symbolic verification technique. In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.

@@

Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).

@@

To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}. This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap. A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.\ pointers that cannot alias). We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in \S\ref{corral}.

@@

\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise}). Such sets can be typically modeled in Boogie using maps. However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness. To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, which works as follows:

@@

Statically analyzing system-level software, such as drivers, requires to ``close'' the environment by abstracting away the low level implementation details. In this work, we developed a simple environmental model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in decreased precision and, thus, false positives. However, because we currently only focus on finding data races, we can get away with over-approximating a lot of the underlying Linux kernel functionality, without losing too much precision. Enriching our model with domain-specific knowledge to make it more precise is an ongoing manual effort, but requires Linux expertise. We argue that further increasing the precision is orthogonal to the contributions of this paper. Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.
